Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-08-10 chapters/Paper_one.tex
All references polished, conclusion missing

\section{Introduction}
Thin layers of liquids on solid surfaces are frequently encountered in a host
of natural and technological settings \cite{degennesCapillarityWettingPhenomena2004,Focke}. Therefore,
understanding and controlling their stability and dynamics is a central problem
for fundamental physics, as well as for applied research in process engineering
and nanotechnology~\cite{oronLongscaleEvolutionThin1997,Utada}. Coating processes, for instance,
rely crucially on the mutual affinity of liquid and surface (i.e. on
wettability properties). When the liquid film is sufficiently thin, in fact, it
can become unstable, leading to the dewetting of the coated
area~\cite{bonnWettingSpreading2009}. From the modelling point of view, the challenge
consists in the fact that the physics of thin films is intrinsically
multiscale, for it involves phenomena ranging from the molecular scale 
at the three phase contact line, to the micro-/nano-metric size of the 
film thickness, to the size of the film as a whole, extending over the coated substrate area.

A fully resolved bottom-up atomistic
approach would be, obviously, unfeasible, if hydrodynamic regimes are to be
explored. It clearly appears that some degree of model order reduction is required. Most hydrodynamic models of thin liquid films, in the framework of the
lubrication theory, simplify the complexity of the full 3D Navier-Stokes
equations \cite{Navier,Stokes}
to one scalar transport equation (the lubrication equation) for the film thickness field
$h(x,t)$~\cite{reynoldsTheoryLubricationIts1886,oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009,Mitlin}:
\begin{equation}\label{eq:lubr}
    \partial_t h = \nabla \cdot \left(Q(h)\nabla p_{\mbox{\tiny{film}}}\right)
\end{equation}
Here, $Q(h)$ is the mobility function, whose explicit form depends on the
boundary condition for the velocity at the surface (for a no-slip boundary, $Q(h) = h^3/(3 \mu)$, with
$\mu$ being the dynamic viscosity), and $p_{\mbox{\tiny{film}}}$ is the film
pressure at the free liquid surface.
Stable and reliable direct numerical simulations of Eq.~(\ref{eq:lubr}) require sophisticated numerical methods, whose execution is often computationally expensive~\cite{beckerComplexDewettingScenarios2003}.
Moreover, an ever-growing number of microfluidic problems requires to cope with
complex fluids rather than simple liquids, i.e. fluids with non-trivial
internal microstructure and/or complex non-Newtonian rheological
behaviour (e.g. colloidal suspensions, polymer solutions, etc.). 
The quest for an efficient multiscale numerical
method for simulating thin film hydrodynamics, versatile for the inclusion of
multiphysics features, is, thus, an ongoing endeavour.

In this paper, we
present a novel approach to the numerical study of thin liquid films, based on
the lattice Boltzmann method (LBM) \cite{succi}. Due to the
built-in properties of the LBM, our method enjoys an outstanding
computational performance, especially on parallel architectures and graphics
processing units.

@@

When a layer of fluid is characterized by a vertical length scale $H$ much
smaller than the longitudinal one $L$, the equations of motion can be
simplified under the approximation that the ratio of the length scales, 
$\varepsilon \equiv H/L$, is small ($\varepsilon \ll1$, see Fig.~\ref{fig:scheme}).
\begin{figure}
    \centering
    \includegraphics[width = 0.75\textwidth]{graphics/Fig_1_Scheme_V3_filters_2.png}
    \caption{Schematic sketch of a model system: a thin liquid film deposited on a flat substrate. The air-liquid interface is represented by the \textit{height} $h(x,y,t)$. The characteristic thickness of the film is given by $H$.}
    \label{fig:scheme}
\end{figure}
In this limit, and for small reduced Reynolds number, $\varepsilon^2Re$
the lubrication approximation tells that the dynamics is governed by equation (\ref{eq:lubr}).
Instead of directly solving Eq.~(\ref{eq:lubr}) numerically, we follow an alternative strategy.
We build our numerical model on a class of LBMs originally proposed as solvers for
the shallow water
equations~\cite{salmonLatticeBoltzmannMethod1999,dellarNonhydrodynamicModesPriori2002,zhouLatticeBoltzmannMethods2004,vanthangStudy1DLattice2010}.
The lattice Boltzmann equation for the discrete probability density functions of a fluid system
subject to a total force (that can include both internal and external forces)
$\mathbf{F}_{\mbox{\tiny{tot}}}$,
$f_l(\mathbf{x},t)$, reads:
\begin{equation}\label{eq:LBE}
\begin{split}
&f_l(\mathbf{x}+\mathbf{c}^{(l)}\Delta t,t+\Delta t) = \\
&(1 - \omega) f_l(\mathbf{x},t) + \omega f_l^{(eq)}(\mathbf{x},t) + w_l \frac{\Delta t}{c_s^2} \mathbf{c}^{(l)} \cdot \mathbf{F}_{\mbox{\tiny{tot}}},
\end{split}
\end{equation}
where $l$ labels the lattice velocities $\mathbf{c}_l$ and runs from $0$ to $Q-1$, with $Q$ being the number
of velocities characterizing the scheme.
Algorithmically, this equation can be seen as made up of two steps. A \textit{local collision}
step where the $f_l(\mathbf{x},t)$ ``relax'' towards the local equilibrium distributions
$f^{(eq)}_l(\mathbf{x},t)$ with rate  
$\omega = \Delta t/\tau$ (where $\tau$, the relaxation time, is proportional to the kinematic viscosity $\nu$): the distribution functions are substituted by their weighted average 
(with weights $\omega$ and $1-\omega$) 
with the equilibria, with an added so-called "source" term (the last term on the right hand side of Eq. (\ref{eq:LBE})), when a force is present. A \textit{non-local streaming} step where the updated distribution functions are scattered to the nearest neighbouring sites. 
The parameters $c_s$ (the lattice speed of sound) and $w_l$ (the 
so called "weights")
depend on the geometry of the lattice and are determined under suitable constraints on 
the form of the tensorial moments in the lattice velocities up to fourth order~\cite{wolf-gladrow}.
We work with two-dimensional square lattices of side length $N\Delta x$, with lattice constant $\Delta x$ and $Q=9$.
For simplicity, we keep $\Delta t=\Delta x=1$ throughout this paper and follow the standard notation, where $c_s = 1/\sqrt{3}$ and the $\mathbf{c}^{(l)} = (c^{(l)}_x,c^{(l)}_y)$, $l=0,1,\dots,8$, are~\cite{qian-dhumieres-lallemand,shan_yuan_chen_2006}

@@

The equilibrium distribution functions $f_l^{(eq)}$ have to be determined to recover the
desired equations of motion for hydrodynamic fields in the long wavelength limit
(we will return to this shortly). They have, therefore, to fulfill the following 
relations involving the liquid height

@@

where the left hand side coincides with the momentum flux of the shallow water equation, with the term 
$gh^2/2$ being the hydrostatic pressure in a thin fluid 
layer at rest~\cite{dellarNonhydrodynamicModesPriori2002}.
With the usual \textit{ansatz}
of a quadratic polynomial in the velocity field $\mathbf{u}$,
the equilibrium distribution functions read

@@

The multiscale Chapman-Enskog expansion \cite{chapmanMathematicalTheoryNonuniform1990,enskogKinetischeTheorieVorgange1917} of such a
LBM yields (for small ratios $Ma/Fr$
of the Mach, $Ma=u/c_s$, and Froude, $Fr = u/\sqrt{gH}$, numbers,
corresponding also to $\sqrt{g H}/c_s \ll 1$) the following equations
for the height and velocity fields~\cite{dellarNonhydrodynamicModesPriori2002,vanthangStudy1DLattice2010,salmonLatticeBoltzmannMethod1999}

@@

where $\nu$, the kinematic viscosity, is related to the relaxation rate $\omega$ appearing
in (\ref{eq:LBE}) via $\nu = c_s^2((2-\omega)/2\omega)\Delta t$. 
For stability of the scheme, the condition $Fr < 1$
is also required, which is fulfilled in all our applications,
given the low values of $u$ (as discussed in more detail later on).
Different terms contribute to the total (generalized) force\footnote{The generalized forces have
indeed the dimensions of $[\mbox{length}]^2[\mbox{time}]^{-2}$.}
$\mathbf{F}_{\mbox{\tiny{tot}}}$:

@@

and $\rho_0$ is the (constant) liquid density 
(equal to $1$, in LBM units). 
The first term in Eq.~(\ref{eq:pfilm}) represents the capillary Laplace pressure (with $\gamma$
being the surface tension) while the second term is the disjoining pressure,
Various forms have been proposed for $\Pi(h)$ in the
literature~\cite{oronLongscaleEvolutionThin1997,thielePatternedDepositionMoving2014}, where here we use the expression

@@

In Eq.~(\ref{eq:disjoiningP}), $\theta$ is the contact angle and $h_*$
corresponds to the precursor film
thickness. The integers $n$ and $m$ are set to be $3$ and $9$, respectively. These 
are commonly chosen values in the literature~\cite{moultonEffectDisjoiningPressure2013,oronLongscaleEvolutionThin1997}
that correspond to a standard $6-12$ Lennard-Jones
intermolecular potential~\cite{fischer2018existence}, though other pairs
$(n,m)$ can be used (e.g.~$(2,3), (3,6), (4,10)$~\cite{1742-6596-166-1-012009,
wilczekSlidingDropsEnsemble2017, wedershoven2014infrared}). 

@@

The film pressure is specific to model thin film dynamics, in general however one can make use of other force terms e.g. to couple fluid layers which has been shown in \cite{doi:10.1002/fld.2742}. The second term on the right hand side in Eq.~(\ref{eq:force}) introduces
a friction with the substrate of the form

@@

Here, $\delta$ acts as a regularizing parameter, which can 
be identified with an effective slip length. 
Finally, the last term in Eq.~(\ref{eq:force}), $\mathbf{F}$ accounts for any other possible source
of forcing (e.g. the gravity component parallel to the substrate in the case of a liquid film
deposited on an inclined plate).

@@

Let us notice at this point that for most microfluidic applications we are actually
interested in, the advection term on the left hand side of the second equation of
(\ref{eq:hydro2}) is indeed negligible as compared to the right hand side
(the Reynolds number \cite{Reynolds,Sommerfeld} being much smaller than one).  
Analogously, the longitudinal viscous terms $\nu \nabla^2 (h\mathbf{u})$ and
$2\nu \nabla (\nabla \cdot (h\mathbf{u}))$ are of order $\varepsilon^2$ smaller 
in the ratio of length scales than the friction term $\nu \alpha(h) \mathbf{u}$
(since the former scale as $\nu H\frac{U}{L^2}$, whereas the latter as $\nu \frac{U}{H}$).

@@

The validity of these considerations has been numerically tested in some selected cases
(representative of typical applications) and the results are shown and discussed in the appendix.

@@

For processes evolving on time-scales $\tilde{t}$ such that $\tilde{t}\gg \frac{h}{\alpha(h)\nu}$, one 
can consider the ``quasi-steady'' limit of the second of these equations 
(setting $\partial_t (h\mathbf{u}) \approx 0$), which yields

@@

In the no slip limit, $\delta \rightarrow 0$, and in absence of gravity and other 
forces, Eq.~(\ref{eq:uslaved}) simplifies into

@@

with the dynamic viscosity $\mu = \rho_0 \nu$. Inserting this result into the first equation of (\ref{eq:hydro3}) leads to

@@

In essence, our method is, therefore, an alternative solver of the lubrication equation (at least in the inertialess regime, $Re \ll 1$, 
and for very thin films, $\varepsilon \ll 1$), that brings in, from the computational point of view, the added values of excellent scalability of the
corresponding LBM algorithm on parallel architectures, as we shall see in the
following sections. Similar ideas have also been developed for 
reaction-diffusion equations~\cite{PhysRevA.45.5771,Kingdon_1992,CHEN1995617,WEIMAR1996207} 
and the modelling of surface tension effects by gradients of auxiliary fields~\cite{PhysRevE.54.5041, PhysRevE.67.036701} based on the color gradient method~\cite{PhysRevA.43.4320}.

@@

The Rayleigh-Taylor instability occurs when a denser fluid is accelerated 
against a less dense one~\cite{Rayleigh,Taylor,KULL1991197,Sharp}. This can be the case, for instance, 
for a liquid film coating a ceiling, under the action of gravity. In such a configuration
gravity tends, of course, to deform (and eventually disrupt) the film, while surface tension has a stabilizing effect.
As a result of these competing mechanisms, any surface perturbation is stable or unstable
depending on whether its characteristic wavenumber $k$ is smaller or larger than a certain
critical value $k_c$. Linear stability analysis calculations in the framework of lubrication theory provide the following growth rate $\sigma(k)$:
\begin{equation}\label{eq:RTgrowth}
    \sigma(k) = \frac{\rho g h_0^3}{3\mu}(k^2 - l_{\mbox{\tiny{cap}}}^2 k^4),
\end{equation}
where $l_{\mbox{\tiny{cap}}} = (\gamma/g)^{1/2}$ is the capillary length. 
Unstable (stable) modes correspond to
$\sigma(k)>0$ ($\sigma(k)<0$) and the critical wavenumber is, therefore, such that $\sigma(k_c)=0$, i.e. 
$k_c = 1/l_{\mbox{\tiny{cap}}}$.
On a lattice of size $2048 \times 2048$ nodes, we initialize the film height according to
\begin{equation}
 h(\mathbf{x},0) = h_0(1 + \varepsilon(\mathbf{x})),
\end{equation}
with $\varepsilon$ a random variable homogeneously distributed in $[1\cdot 10^{-4},-1\cdot 10^{-4}]$ and $h_0 = 1$. Forcing should always be below a certain threshold. Thus, for the gravitational acceleration we choose values within the interval $|g| = [4,8]\cdot 10^{-5}$. Furthermore, we fix the surface tension to be $\gamma=0.01$. This results in critical wavenumbers ranging from $k_c= 0.06$ to $k_c = 0.09$.
Fig.~\ref{fig:RTI_evolution} shows snapshots of the free surface from various time steps, where the growth of the perturbations is shown as time increases. The last panel is already beyond the linear regime.

@@

\begin{figure}
    \centering 
    \includegraphics[width=0.75\textwidth]{graphics/Fig_5_Indirect_Cox_Voinov_all_data_visually_appealing_slip_2_m_nosci.png}
    \caption{Difference of cubed instantaneous and equilibrium contact angles, $\theta_{num}^3-\theta_{eq}^3$, vs. capillary number $Ca$ for a spreading droplet; the dashed line shows a linear dependence (consistent with the Cox-Voinov law). The different symbols represent different viscosities, while the dashed line is a linear function of the capillary number.
    \label{fig:Cox-Voinov}}
\end{figure}
The spreading dynamics can be investigated even more quantitatively in terms of the 
so-called Cox-Voinov law and Tanner's law~\cite{Tanner_1979}. The first one relates the apparent contact angle to 
the velocity $U$ of the spreading front (the contact line), at various times, by 
$\theta^3 - \theta_{eq}^3 \propto Ca$. The capillary number $Ca$ is defined
as $Ca=\mu U/\gamma$~\cite{doi:10.1146/annurev-fluid-011212-140734}.
In Fig.~\ref{fig:Cox-Voinov} we plot $\theta^3(t) - \theta_{eq}^3$ vs $Ca(t)$ from a numerical simulation of a spreading drop: a good linear scaling, in agreement with the Cox-Voinov law, is 
observed, as highlighted by the dashed line.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_6_Tanners_law_slip_2_paper_rescaled_t.png}
    \caption{Time evolution of the droplet base radius of a spreading droplet; the dashed red line shows a $\tilde{t}^{1/10}$ power law (In consistence with Tanner's law). As in Fig.~\ref{fig:Cox-Voinov} different symbols refer to different viscosities. The radius clearly grows with the predicted power law until it saturates. Upon rescaling the time with $\tau_{\mbox{\tiny{cap}}}$ the curves of all three viscosities collapse into a single one.}
    \label{fig:Tanners_law}
\end{figure}
The Tanner's law which states that the radius of the droplet grows with time as
\begin{equation}
\label{eq:tanners_law}
    R(t)\approx \left[\frac{10\gamma}{9B\mu}\left(\frac{4V}{\pi}\right)^3 t\right]^{1/10},
\end{equation}
with the constant $B$ being such that 
$B^{1/10} \approx 1.2$.  
In Fig.~\ref{fig:Tanners_law} 
we plot the measured radius of the droplet divided by its initial radius $R_0$ as a function of the dimensionless time $\tilde{t} = t/\tau_{\mbox{\tiny{cap}}}$ 
(here $\tau_{\mbox{\tiny{cap}}} = \frac{\mu R}{\gamma}$). For the three viscosities considered 
in Figs.~\ref{fig:Cox-Voinov},\ref{fig:Tanners_law} our capillary times are 
$\tau_{\mbox{\tiny{cap}}} = [1333, 1667, 2333] \Delta t$. 
We see a saturation at $R/R_0=1.17$ because the droplet is very close to 
its equilibrium shape. During the growth phase the radius follows indeed a power law 
in $\tilde{t}$ with exponent $1/10$, which is shown by the red dashed line, in agreement with Tanner's prediction and 
experimental results \cite{rioboo2002time, jambon-puillet_carrier_shahidzadeh_brutin_eggers_bonn_2018, cazabat1986dynamics, CHEN198860}.
We further notice that within our
simulations the droplet needs about $12\tau_{\mbox{\tiny{cap}}}$ to reach its equilibrium shape.

>>>>

\section{Introduction}
Thin layers of liquids on solid surfaces are frequently encountered in a host of natural and technological settings \cite{degennesCapillarityWettingPhenomena2004,fockeLabonaFoilMicrofluidicsThin2010}. 
Therefore, understanding and controlling their stability and dynamics is a central problem for fundamental physics, as well as for applied research in process engineering and nanotechnology~\cite{oronLongscaleEvolutionThin1997,utadaDrippingJettingDrops2007}. 
Coating processes, for instance, rely crucially on the mutual affinity of liquid and surface (i.e. on wettability properties). When the liquid film is sufficiently thin, in fact, it can become unstable, leading to the dewetting of the coated area~\cite{bonnWettingSpreading2009}. 
From the modelling point of view, the challenge consists in the fact that the physics of thin films is intrinsically multiscale, for it involves phenomena ranging from the molecular scale at the three phase contact line, to the micro-/nano-metric size of the film thickness, to the size of the film as a whole, extending over the coated substrate area.

A fully resolved bottom-up atomistic approach would be, obviously, unfeasible, if hydrodynamic regimes are to be explored. 
It clearly appears that some degree of model order reduction is required. 
Most hydrodynamic models of thin liquid films, in the framework of the lubrication theory, simplify the complexity of the full 3D Navier-Stokes equations~\cite{claude-louis-marie-henriMemoireLoisMouvement1827, stokesSteadyMotionIncompressible1848} to one scalar transport equation (the lubrication equation) for the film thickness field $h(x,t)$~\cite{reynoldsTheoryLubricationIts1886,oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009,mitlinDewettingSolidSurface1993}:
\begin{equation}\label{eq:lubr}
    \partial_t h = \nabla \cdot \left(Q(h)\nabla p_{\mbox{\tiny{film}}}\right)
\end{equation}
Here, $Q(h)$ is the mobility function, whose explicit form depends on the boundary condition for the velocity at the surface (for a no-slip boundary, $Q(h) = h^3/(3 \mu)$, with $\mu$ being the dynamic viscosity), and $p_{\mbox{\tiny{film}}}$ is the film pressure at the free liquid surface.
Stable and reliable direct numerical simulations of Eq.~(\ref{eq:lubr}) require sophisticated numerical methods, whose execution is often computationally expensive~\cite{beckerComplexDewettingScenarios2003}.
Moreover, an ever-growing number of microfluidic problems requires to cope with complex fluids rather than simple liquids, i.e. fluids with non-trivial internal microstructure and/or complex non-Newtonian rheological behaviour (e.g. colloidal suspensions, polymer solutions, etc.). 
The quest for an efficient multiscale numerical method for simulating thin film hydrodynamics, versatile for the inclusion of multiphysics features, is, thus, an ongoing endeavour.

In this paper, we present a novel approach to the numerical study of thin liquid films, based on the lattice Boltzmann method (LBM) \cite{succiLatticeBoltzmannEquation2001}. 
Due to the built-in properties of the LBM, our method enjoys an outstanding computational performance, especially on parallel architectures and graphics processing units.

@@

When a layer of fluid is characterized by a vertical length scale $H$ much smaller than the longitudinal one $L$, the equations of motion can be simplified under the approximation that the ratio of the length scales, $\varepsilon \equiv H/L$, is small ($\varepsilon \ll1$, see Fig.~\ref{fig:scheme}).
\begin{figure}
    \centering
    \includegraphics[width = 0.75\textwidth]{graphics/Fig_1_Scheme_V3_filters_2.png}
    \caption{Schematic sketch of a model system: a thin liquid film deposited on a flat substrate. The air-liquid interface is represented by the \textit{height} $h(x,y,t)$. The characteristic thickness of the film is given by $H$.}
    \label{fig:scheme}
\end{figure}
In this limit, and for small reduced Reynolds number, $\varepsilon^2Re$ the lubrication approximation tells that the dynamics is governed by equation (\ref{eq:lubr}).
Instead of directly solving Eq.~(\ref{eq:lubr}) numerically, we follow an alternative strategy.
We build our numerical model on a class of LBMs originally proposed as solvers for the shallow water equations~\cite{salmonLatticeBoltzmannMethod1999,dellarNonhydrodynamicModesPriori2002,zhouLatticeBoltzmannMethods2004,vanthangStudy1DLattice2010}.
The lattice Boltzmann equation for the discrete probability density functions of a fluid system subject to a total force (that can include both internal and external forces) $\mathbf{F}_{\mbox{\tiny{tot}}}$, $f_l(\mathbf{x},t)$, reads:
\begin{equation}\label{eq:LBE}
\begin{split}
&f_l(\mathbf{x}+\mathbf{c}^{(l)}\Delta t,t+\Delta t) = \\
&(1 - \omega) f_l(\mathbf{x},t) + \omega f_l^{(eq)}(\mathbf{x},t) + w_l \frac{\Delta t}{c_s^2} \mathbf{c}^{(l)} \cdot \mathbf{F}_{\mbox{\tiny{tot}}},
\end{split}
\end{equation}
where $l$ labels the lattice velocities $\mathbf{c}_l$ and runs from $0$ to $Q-1$, with $Q$ being the number of velocities characterizing the scheme.
Algorithmically, this equation can be seen as made up of two steps. 
A \textit{local collision} step where the $f_l(\mathbf{x},t)$ ``relax'' towards the local equilibrium distributions $f^{(eq)}_l(\mathbf{x},t)$ with rate $\omega = \Delta t/\tau$ (where $\tau$, the relaxation time, is proportional to the kinematic viscosity $\nu$): the distribution functions are substituted by their weighted average (with weights $\omega$ and $1-\omega$) with the equilibria, with an added so-called "source" term (the last term on the right hand side of Eq. (\ref{eq:LBE})), when a force is present. 
A \textit{non-local streaming} step where the updated distribution functions are scattered to the nearest neighbouring sites. 
The parameters $c_s$ (the lattice speed of sound) and $w_l$ (the so called "weights") depend on the geometry of the lattice and are determined under suitable constraints on the form of the tensorial moments in the lattice velocities up to fourth order~\cite{wolf-gladrowLatticeGasCellularAutomata2004}.
We work with two-dimensional square lattices of side length $N\Delta x$, with lattice constant $\Delta x$ and $Q=9$.
For simplicity, we keep $\Delta t=\Delta x=1$ throughout this paper and follow the standard notation, where $c_s = 1/\sqrt{3}$ and the $\mathbf{c}^{(l)} = (c^{(l)}_x,c^{(l)}_y)$, $l=0,1,\dots,8$, are~\cite{qianFractionalPropagationElimination1997,shanLatticeBoltzmannModel1993}

@@

The equilibrium distribution functions $f_l^{(eq)}$ have to be determined to recover the desired equations of motion for hydrodynamic fields in the long wavelength limit (we will return to this shortly). 
They have, therefore, to fulfill the following relations involving the liquid height

@@

where the left hand side coincides with the momentum flux of the shallow water equation, with the term $gh^2/2$ being the hydrostatic pressure in a thin fluid layer at rest~\cite{dellarNonhydrodynamicModesPriori2002}.
With the usual \textit{ansatz} of a quadratic polynomial in the velocity field $\mathbf{u}$, the equilibrium distribution functions read

@@

The multiscale Chapman-Enskog expansion \cite{chapmanMathematicalTheoryNonuniform1990,enskogKinetischeTheorieVorgange1917} of such a LBM yields (for small ratios $Ma/Fr$ of the Mach, $Ma=u/c_s$, and Froude, $Fr = u/\sqrt{gH}$, numbers, corresponding also to $\sqrt{g H}/c_s \ll 1$) the following equations for the height and velocity fields~\cite{dellarNonhydrodynamicModesPriori2002,vanthangStudy1DLattice2010,salmonLatticeBoltzmannMethod1999}

@@

where $\nu$, the kinematic viscosity, is related to the relaxation rate $\omega$ appearing in (\ref{eq:LBE}) via $\nu = c_s^2((2-\omega)/2\omega)\Delta t$. 
For stability of the scheme, the condition $Fr < 1$ is also required, which is fulfilled in all our applications, given the low values of $u$ (as discussed in more detail later on).
Different terms contribute to the total (generalized) force\footnote{The generalized forces have indeed the dimensions of $[\mbox{length}]^2[\mbox{time}]^{-2}$.} $\mathbf{F}_{\mbox{\tiny{tot}}}$:

@@

and $\rho_0$ is the (constant) liquid density (equal to $1$, in LBM units). 
The first term in Eq.~(\ref{eq:pfilm}) represents the capillary Laplace pressure (with $\gamma$ being the surface tension) while the second term is the disjoining pressure, 
Various forms have been proposed for $\Pi(h)$ in the literature~\cite{oronLongscaleEvolutionThin1997,thielePatternedDepositionMoving2014}, where here we use the expression

@@

In Eq.~(\ref{eq:disjoiningP}), $\theta$ is the contact angle and $h_*$ corresponds to the precursor film thickness. 
The integers $n$ and $m$ are set to be $3$ and $9$, respectively. 
These are commonly chosen values in the literature~\cite{moultonEffectDisjoiningPressure2013,oronLongscaleEvolutionThin1997} that correspond to a standard $6-12$ Lennard-Jones intermolecular potential~\cite{fischerExistencePositiveSolutions2018}, though other pairs $(n,m)$ can be used (e.g.~$(2,3), (3,6), (4,10)$~\cite{diezStabilityFinitelengthRivulet2009,wilczekSlidingDropsEnsemble2017, wedershovenInfraredLaserInduced2014}). 

@@

The film pressure is specific to model thin film dynamics, in general however one can make use of other force terms e.g. to couple fluid layers which has been shown in \cite{roccaDevelopmentLatticeBoltzmann2012}. 
The second term on the right hand side in Eq.~(\ref{eq:force}) introduces a friction with the substrate of the form

@@

Here, $\delta$ acts as a regularizing parameter, which can be identified with an effective slip length. 
Finally, the last term in Eq.~(\ref{eq:force}), $\mathbf{F}$ accounts for any other possible source of forcing (e.g. the gravity component parallel to the substrate in the case of a liquid film deposited on an inclined plate).

@@

Let us notice at this point that for most microfluidic applications we are actually interested in, the advection term on the left hand side of the second equation of (\ref{eq:hydro2}) is indeed negligible as compared to the right hand side (the Reynolds number \cite{reynoldsIIIExperimentalInvestigation1883,sommerfeldBeitrag1909} being much smaller than one).  
Analogously, the longitudinal viscous terms $\nu \nabla^2 (h\mathbf{u})$ and $2\nu \nabla (\nabla \cdot (h\mathbf{u}))$ are of order $\varepsilon^2$ smaller in the ratio of length scales than the friction term $\nu \alpha(h) \mathbf{u}$ (since the former scale as $\nu H\frac{U}{L^2}$, whereas the latter as $\nu \frac{U}{H}$).

@@

The validity of these considerations has been numerically tested in some selected cases (representative of typical applications) and the results are shown and discussed in the appendix.

@@

For processes evolving on time-scales $\tilde{t}$ such that $\tilde{t}\gg \frac{h}{\alpha(h)\nu}$, one can consider the ``quasi-steady'' limit of the second of these equations (setting $\partial_t (h\mathbf{u}) \approx 0$), which yields

@@

In the no slip limit, $\delta \rightarrow 0$, and in absence of gravity and other forces, Eq.~(\ref{eq:uslaved}) simplifies into

@@

with the dynamic viscosity $\mu = \rho_0 \nu$. 
Inserting this result into the first equation of (\ref{eq:hydro3}) leads to

@@

In essence, our method is, therefore, an alternative solver of the lubrication equation (at least in the inertialess regime, $Re \ll 1$, and for very thin films, $\varepsilon \ll 1$), that brings in, from the computational point of view, the added values of excellent scalability of the corresponding LBM algorithm on parallel architectures, as we shall see in the following sections. 
Similar ideas have also been developed for reaction-diffusion equations~\cite{caliDiffusionHydrodynamicDispersion1992,kingdonReactionflowLatticeBoltzmann1992,chenLatticeMethodsTheir1995,weimarNonlinearReactionsAdvected1996} and the modelling of surface tension effects by gradients of auxiliary fields~\cite{swiftLatticeBoltzmannSimulations1996, lishchukLatticeBoltzmannAlgorithm2003} based on the color gradient method~\cite{gunstensenLatticeBoltzmannModel1991}.

@@

The Rayleigh-Taylor instability occurs when a denser fluid is accelerated against a less dense one~\cite{rayleighInvestigationCharacterEquilibrium1882,taylorInstabilityLiquidSurfaces1997,kullTheoryRayleighTaylorInstability1991,sharpOverviewRayleighTaylorInstability1984}. 
This can be the case, for instance, for a liquid film coating a ceiling, under the action of gravity. 
In such a configuration gravity tends, of course, to deform (and eventually disrupt) the film, while surface tension has a stabilizing effect.
As a result of these competing mechanisms, any surface perturbation is stable or unstable depending on whether its characteristic wavenumber $k$ is smaller or larger than a certain critical value $k_c$. 
Linear stability analysis calculations in the framework of lubrication theory provide the following growth rate $\sigma(k)$:
\begin{equation}\label{eq:RTgrowth}
    \sigma(k) = \frac{\rho g h_0^3}{3\mu}(k^2 - l_{\mbox{\tiny{cap}}}^2 k^4),
\end{equation}
where $l_{\mbox{\tiny{cap}}} = (\gamma/g)^{1/2}$ is the capillary length. 
Unstable (stable) modes correspond to $\sigma(k)>0$ ($\sigma(k)<0$) and the critical wavenumber is, therefore, such that $\sigma(k_c)=0$, i.e. $k_c = 1/l_{\mbox{\tiny{cap}}}$.
On a lattice of size $2048 \times 2048$ nodes, we initialize the film height according to
\begin{equation}
  h(\mathbf{x},0) = h_0(1 + \varepsilon(\mathbf{x})),
\end{equation}
with $\varepsilon$ a random variable homogeneously distributed in $[1\cdot 10^{-4},-1\cdot 10^{-4}]$ and $h_0 = 1$. 
Forcing should always be below a certain threshold. Thus, for the gravitational acceleration we choose values within the interval $|g| = [4,8]\cdot 10^{-5}$. Furthermore, we fix the surface tension to be $\gamma=0.01$. 
This results in critical wavenumbers ranging from $k_c= 0.06$ to $k_c = 0.09$.
Fig.~\ref{fig:RTI_evolution} shows snapshots of the free surface from various time steps, where the growth of the perturbations is shown as time increases. The last panel is already beyond the linear regime.

@@

\begin{figure}
    \centering 
    \includegraphics[width=0.75\textwidth]{graphics/Fig_5_Indirect_Cox_Voinov_all_data_visually_appealing_slip_2_m_nosci.png}
    \caption{Difference of cubed instantaneous and equilibrium contact angles, $\theta_{num}^3-\theta_{eq}^3$, vs. capillary number $Ca$ for a spreading droplet; the dashed line shows a linear dependence (consistent with the Cox-Voinov law). The different symbols represent different viscosities, while the dashed line is a linear function of the capillary number.
    \label{fig:Cox-Voinov}}
\end{figure}
The spreading dynamics can be investigated even more quantitatively in terms of the so-called Cox-Voinov law and Tanner's law~\cite{tannerSpreadingSiliconeOil1979}. 
The first one relates the apparent contact angle to the velocity $U$ of the spreading front (the contact line), at various times, by $\theta^3 - \theta_{eq}^3 \propto Ca$. 
The capillary number $Ca$ is defined as $Ca=\mu U/\gamma$~\cite{snoeijerMovingContactLines2013}.
In Fig.~\ref{fig:Cox-Voinov} we plot $\theta^3(t) - \theta_{eq}^3$ vs $Ca(t)$ from a numerical simulation of a spreading drop: a good linear scaling, in agreement with the Cox-Voinov law, is observed, as highlighted by the dashed line.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_6_Tanners_law_slip_2_paper_rescaled_t.png}
    \caption{Time evolution of the droplet base radius of a spreading droplet; the dashed red line shows a $\tilde{t}^{1/10}$ power law (In consistence with Tanner's law). As in Fig.~\ref{fig:Cox-Voinov} different symbols refer to different viscosities. The radius clearly grows with the predicted power law until it saturates. Upon rescaling the time with $\tau_{\mbox{\tiny{cap}}}$ the curves of all three viscosities collapse into a single one.}
    \label{fig:Tanners_law}
\end{figure}
The Tanner's law which states that the radius of the droplet grows with time as
\begin{equation}
\label{eq:tanners_law}
    R(t)\approx \left[\frac{10\gamma}{9B\mu}\left(\frac{4V}{\pi}\right)^3 t\right]^{1/10},
\end{equation}
with the constant $B$ being such that $B^{1/10} \approx 1.2$.  
In Fig.~\ref{fig:Tanners_law} we plot the measured radius of the droplet divided by its initial radius $R_0$ as a function of the dimensionless time $\tilde{t} = t/\tau_{\mbox{\tiny{cap}}}$ (here $\tau_{\mbox{\tiny{cap}}} = \frac{\mu R}{\gamma}$). 
For the three viscosities considered in Figs.~\ref{fig:Cox-Voinov},\ref{fig:Tanners_law} our capillary times are $\tau_{\mbox{\tiny{cap}}} = [1333, 1667, 2333] \Delta t$. 
We see a saturation at $R/R_0=1.17$ because the droplet is very close to its equilibrium shape. 
During the growth phase the radius follows indeed a power law in $\tilde{t}$ with exponent $1/10$, which is shown by the red dashed line, in agreement with Tanner's prediction and experimental results \cite{riobooTimeEvolutionLiquid2002, jambon-puilletSpreadingDynamicsContact2018, cazabatDynamicsWettingEffects1986, chenExperimentsSpreadingDrop1988}.
We further notice that within our simulations the droplet needs about $12\tau_{\mbox{\tiny{cap}}}$ to reach its equilibrium shape.
